Reasoning about a Capability Machine with Local Capabilities\\ Provably Safe Stack and Return Pointer Management (without OS Support)\\ Technical Appendix Including Proofs and Details

### 2017-06-24 fully-abstract-comp/technical-report.tex
merge

% Two arguments
\newcommand{\sjnz}[2]{\twoinstr{\src{jnz}}{#1}{#2}}
\newcommand{\sisptr}[2]{\twoinstr{\src{isptr}}{#1}{#2}}
\newcommand{\sgeta}[2]{\twoinstr{\src{geta}}{#1}{#2}}
\newcommand{\sgetb}[2]{\twoinstr{\src{getb}}{#1}{#2}}
\newcommand{\sgete}[2]{\twoinstr{\src{gete}}{#1}{#2}}
\newcommand{\sgetp}[2]{\twoinstr{\src{getp}}{#1}{#2}}
\newcommand{\sgetl}[2]{\twoinstr{\src{getl}}{#1}{#2}}
\newcommand{\smove}[2]{\twoinstr{\src{move}}{#1}{#2}}
\newcommand{\sstore}[2]{\twoinstr{\src{store}}{#1}{#2}}
\newcommand{\sload}[2]{\twoinstr{\src{load}}{#1}{#2}}
\newcommand{\slea}[2]{\twoinstr{\src{lea}}{#1}{#2}}
\newcommand{\ssload}[2]{\twoinstr{\src{sload}}{#1}{#2}}
\newcommand{\scall}[2]{\twoinstr{\src{call}}{#1}{#2}}

% Three arguments
\newcommand{\srestrict}[3]{\threeinstr{\src{restrict}}{#1}{#2}{#3}}
\newcommand{\ssubseg}[3]{\threeinstr{\src{subseg}}{#1}{#2}{#3}}
\newcommand{\splus}[3]{\threeinstr{\src{plus}}{#1}{#2}{#3}}
\newcommand{\sminus}[3]{\threeinstr{\src{minus}}{#1}{#2}{#3}}

@@

% Two arguments
\newcommand{\tjnz}[2]{\twoinstr{\trg{jnz}}{#1}{#2}}
\newcommand{\tisptr}[2]{\twoinstr{\trg{isptr}}{#1}{#2}}
\newcommand{\tgeta}[2]{\twoinstr{\trg{geta}}{#1}{#2}}
\newcommand{\tgetb}[2]{\twoinstr{\trg{getb}}{#1}{#2}}
\newcommand{\tgete}[2]{\twoinstr{\trg{gete}}{#1}{#2}}
\newcommand{\tgetp}[2]{\twoinstr{\trg{getp}}{#1}{#2}}
\newcommand{\tgetl}[2]{\twoinstr{\trg{getl}}{#1}{#2}}
\newcommand{\tmove}[2]{\twoinstr{\trg{move}}{#1}{#2}}
\newcommand{\tstore}[2]{\twoinstr{\trg{store}}{#1}{#2}}
\newcommand{\tload}[2]{\twoinstr{\trg{load}}{#1}{#2}}
\newcommand{\tlea}[2]{\twoinstr{\trg{lea}}{#1}{#2}}
\newcommand{\txjmp}[2]{\twoinstr{\trg{xjmp}}{#1}{#2}}
\newcommand{\tcseal}[2]{\twoinstr{\trg{cseal}}{#1}{#2}}

% Three arguments
\newcommand{\trestrict}[3]{\threeinstr{\trg{restrict}}{#1}{#2}{#3}}
\newcommand{\tsubseg}[3]{\threeinstr{\trg{subseg}}{#1}{#2}{#3}}
\newcommand{\tplus}[3]{\threeinstr{\trg{plus}}{#1}{#2}{#3}}
\newcommand{\tminus}[3]{\threeinstr{\trg{minus}}{#1}{#2}{#3}}

@@

%%% Variables
\newcommand{\var}[1]{\mathit{#1}}
\newcommand{\rn}{\var{rn}}
\newcommand{\reg}{\var{reg}}
\newcommand{\mem}{\var{mem}}
\newcommand{\ms}{\var{ms}}
\newcommand{\pc}{\var{pc}}
\newcommand{\stk}{\var{stk}}
\newcommand{\opc}{\var{opc}}

@@

%%% Functions
\newcommand{\plainfun}[2]{
  \ifthenelse{\equal{#2}{}}
  {\mathit{#1}}
  {\mathit{#1}(#2)}
}
\newcommand{\updPcAddr}[1]{\plainfun{updatePc}{#1}}
\newcommand{\updPcPerm}[1]{\plainfun{updatePcPerm}{#1}}
\newcommand{\clearms}[1]{\plainfun{clear_{ms}}{#1}}
\newcommand{\clearlocreg}[1]{\plainfun{clear_{loc,reg}}{#1}}

@@

\subsection{Syntax}
The target machine is a simple capability machine with memory capabilities, local capabilities and sealed capabilities\footnote{In previous work, we used enter capabilities but for this the complexity introduced by mixing writable and executable memory is difficult to handle} (Inspired by CHERI). The syntax of the instructions of the target machine is defined as follows:
\[
\begin{array}{rcl}
n &\in & \nats \\
\trg{r} &\in &  \tRegName \\
\trg{\rn} &::= &  \trg{r} \mid n \\
\trg{i} &::= & \tfail \mid \thalt \mid \tjmp{\trg{r}} \mid \tjnz{\trg{r}}{\trg{\rn}} \mid \tisptr{\trg{r}}{\trg{r}} \mid \tgeta{\trg{r}}{\trg{r}} \mid \tgetb{\trg{r}}{\trg{r}} \mid \\
      & &  \tgete{\trg{r}}{\trg{r}}\mid \tgetp{\trg{r}}{\trg{r}} \mid \tgetl{\trg{r}}{\trg{r}} \mid \tmove{\trg{r}}{\trg{\rn}} \mid \tstore{\trg{r}}{\trg{r}} \mid\\
      & &  \tload{\trg{r}}{\trg{r}} \mid \tlea{\trg{r}}{\trg{\rn}} \mid \trestrict{\trg{r}}{\trg{r}}{\trg{\rn}} \mid \tsubseg{\trg{r}}{\trg{\rn}}{\trg{\rn}} \mid \\
      & & \tplus{\trg{r}}{\trg{\rn}}{\trg{\rn}} \mid \tminus{\trg{r}}{\trg{\rn}}{\trg{\rn}} \mid \tsetatob{\trg{r}} \mid \txjmp{r}{r} \mid \tcseal{r}{r}
\end{array}
\]

@@

\subsubsection{Helpful functions}
\[
  \updPcAddr \defeq \dots
\]

\[
  \updPcPerm \defeq \dots
\]

\[
  \clearms \defeq \dots
\]

\[
  \clearlocreg \defeq \dots
\]

\subsubsection{Source Language}
\lau{23-03-2017: $\sjmp{}$ and $\scall{}{}$ needs an update.}
\begin{align*}
  \ssem{\sjmp{r}} = & 
    \begin{cases}
      \Phi
      \begin{aligned}[t]
        & \update{\reg.\pcreg}{\opc} \\
        & \update{\smsstk}{\ms_\stk' \uplus \clearms{\ms_\stk}}\\
        & \update{\sastk}{a_\stk'}\\
        & \update{\sstk}{\stk'}
      \end{aligned} & \Phi(r) = \retptr \text{ and }
      \Phi.\sstk = (\opc,\ms_\stk',a_\stk') :: \stk'\\
      & \\
      \updPc{\updPcPerm{\Phi(r)}} & \text{otherwise}
    \end{cases} \\
\end{align*}
\begin{align*}
  \ssem{\scall{r_1}{r_2}} = & 
    \begin{cases}
      (\mem,\reg',\stk',\ms_\stk') & \Phi = (\mem,\reg,\stk,\ms_\stk) \text{ and }\\
      \\
      \failed & \text{otherwise}
    \end{cases}
\end{align*}
\begin{itemize}
\item $\scall{}{}$ does not do argument spilling anymore. We leave it to the caller to push the arguments for spilling to the stack and update the stack pointer accordingly.
\end{itemize}

@@

>>>>

% Two arguments
\newcommand{\sjnz}[2]{\twoinstr{\src{jnz}}{#1}{#2}}
\newcommand{\sisptr}[2]{\twoinstr{\src{isptr}}{#1}{#2}}
\newcommand{\sgeta}[2]{\twoinstr{\src{geta}}{#1}{#2}}
\newcommand{\sgetb}[2]{\twoinstr{\src{getb}}{#1}{#2}}
\newcommand{\sgete}[2]{\twoinstr{\src{gete}}{#1}{#2}}
\newcommand{\sgetp}[2]{\twoinstr{\src{getp}}{#1}{#2}}
\newcommand{\sgetl}[2]{\twoinstr{\src{getl}}{#1}{#2}}
\newcommand{\smove}[2]{\twoinstr{\src{move}}{#1}{#2}}
\newcommand{\sstore}[2]{\twoinstr{\src{store}}{#1}{#2}}
\newcommand{\sload}[2]{\twoinstr{\src{load}}{#1}{#2}}
\newcommand{\slea}[2]{\twoinstr{\src{lea}}{#1}{#2}}
\newcommand{\ssload}[2]{\twoinstr{\src{sload}}{#1}{#2}}
\newcommand{\scall}[2]{\twoinstr{\src{call}}{#1}{#2}}
\newcommand{\sxjmp}[2]{\twoinstr{\trg{xjmp}}{#1}{#2}}

% Three arguments
\newcommand{\srestrict}[3]{\threeinstr{\src{restrict}}{#1}{#2}{#3}}
\newcommand{\ssubseg}[3]{\threeinstr{\src{subseg}}{#1}{#2}{#3}}
\newcommand{\splus}[3]{\threeinstr{\src{plus}}{#1}{#2}{#3}}
\newcommand{\sminus}[3]{\threeinstr{\src{minus}}{#1}{#2}{#3}}
\newcommand{\scseal}[3]{\threeinstr{\trg{cseal}}{#1}{#2}{#3}}

@@

% Two arguments
\newcommand{\tjnz}[2]{\twoinstr{\trg{jnz}}{#1}{#2}}
\newcommand{\tisptr}[2]{\twoinstr{\trg{isptr}}{#1}{#2}}
\newcommand{\tgeta}[2]{\twoinstr{\trg{geta}}{#1}{#2}}
\newcommand{\tgetb}[2]{\twoinstr{\trg{getb}}{#1}{#2}}
\newcommand{\tgete}[2]{\twoinstr{\trg{gete}}{#1}{#2}}
\newcommand{\tgetp}[2]{\twoinstr{\trg{getp}}{#1}{#2}}
\newcommand{\tgetl}[2]{\twoinstr{\trg{getl}}{#1}{#2}}
\newcommand{\tmove}[2]{\twoinstr{\trg{move}}{#1}{#2}}
\newcommand{\tstore}[2]{\twoinstr{\trg{store}}{#1}{#2}}
\newcommand{\tload}[2]{\twoinstr{\trg{load}}{#1}{#2}}
\newcommand{\tlea}[2]{\twoinstr{\trg{lea}}{#1}{#2}}
\newcommand{\txjmp}[2]{\twoinstr{\trg{xjmp}}{#1}{#2}}

% Three arguments
\newcommand{\trestrict}[3]{\threeinstr{\trg{restrict}}{#1}{#2}{#3}}
\newcommand{\tsubseg}[3]{\threeinstr{\trg{subseg}}{#1}{#2}{#3}}
\newcommand{\tplus}[3]{\threeinstr{\trg{plus}}{#1}{#2}{#3}}
\newcommand{\tminus}[3]{\threeinstr{\trg{minus}}{#1}{#2}{#3}}
\newcommand{\tcseal}[3]{\threeinstr{\trg{cseal}}{#1}{#2}{#3}}

@@

%%% Variables
\newcommand{\var}[1]{\mathit{#1}}
\newcommand{\rn}{\var{rn}}
\newcommand{\reg}{\var{reg}}
\newcommand{\mem}{\var{mem}}
\newcommand{\ms}{\var{ms}}
\newcommand{\pc}{\var{pc}}
\newcommand{\stk}{\var{stk}}
\newcommand{\opc}{\var{opc}}
\newcommand{\vsc}{\var{sc}}
\newcommand{\cb}{\var{cb}}

@@

%%% Functions
\newcommand{\plainfun}[2]{
  \ifthenelse{\equal{#2}{}}
  {\mathit{#1}}
  {\mathit{#1}(#2)}
}
\newcommand{\updPcAddr}[1]{\plainfun{updatePc}{#1}}
\newcommand{\updPcPerm}[1]{\plainfun{updatePcPerm}{#1}}
\newcommand{\nonexec}[1]{\plainfun{nonExecutable}{#1}}
\newcommand{\noRetStkReg}[1]{\plainfun{noRetStk_{reg}}{#1}}
\newcommand{\noRetStkMs}[1]{\plainfun{noRetStk_{ms}}{#1}}

@@

\subsection{Syntax}
The target machine is a simple capability machine with memory capabilities, local capabilities and sealed capabilities\footnote{In previous work, we used enter capabilities but for this the complexity introduced by mixing writable and executable memory is difficult to handle} (Inspired by CHERI). The syntax of the instructions of the target machine is defined as follows:
\[
\begin{array}{rcl}
n &\in & \nats \\
\trg{r} &\in &  \tRegName \\
\trg{\rn} &::= &  \trg{r} \mid n \\
\trg{i} &::= & \tfail \mid \thalt \mid \tjmp{\trg{r}} \mid \tjnz{\trg{r}}{\trg{\rn}} \mid \tisptr{\trg{r}}{\trg{r}} \mid \tgeta{\trg{r}}{\trg{r}} \mid \tgetb{\trg{r}}{\trg{r}} \mid \\
      & &  \tgete{\trg{r}}{\trg{r}}\mid \tgetp{\trg{r}}{\trg{r}} \mid \tgetl{\trg{r}}{\trg{r}} \mid \tmove{\trg{r}}{\trg{\rn}} \mid \tstore{\trg{r}}{\trg{r}} \mid\\
      & &  \tload{\trg{r}}{\trg{r}} \mid \tlea{\trg{r}}{\trg{\rn}} \mid \trestrict{\trg{r}}{\trg{r}}{\trg{\rn}} \mid \tsubseg{\trg{r}}{\trg{\rn}}{\trg{\rn}} \mid \\
      & & \tplus{\trg{r}}{\trg{\rn}}{\trg{\rn}} \mid \tminus{\trg{r}}{\trg{\rn}}{\trg{\rn}} \mid \tsetatob{\trg{r}} \mid \txjmp{r}{r} \mid \tcseal{r}{r}{r}
\end{array}
\]

@@

\subsubsection{Helpful functions}
\[
  \updPcAddr{} \defeq \dots
\]

\[
  \updPcPerm{} \defeq \dots
\]

\[
  \nonexec{} \defeq \dots
\]

\[
  \noRetStkReg{} \defeq \dots %remember to ignore certain registers
\]

\[
  \noRetStkMs{} \defeq \dots %remember to ignore certain registers
\]

\subsubsection{Source Language}
\lau{23-03-2017: $\sjmp{}$ and $\scall{}{}$ needs an update.}
\begin{align*}
  \ssem{\scall{r_1}{r_2}} = & 
    \begin{cases}
      (\mem,\reg',\stk',\ms_\stk') & \Phi = (\mem,\reg,\stk,\ms_\stk) \text{ and } \dots\\
      \\
      \failed & \text{otherwise}
    \end{cases}
\end{align*}

The $\scseal{}{}{}$ instruction seals a sealable capability with a seal.
\begin{align*}
  \ssem{\scseal{r_1}{r_2}{r_3}} = & 
                                    \begin{cases}
                                      \updReg{r_1}{\vsc} & 
                                      \arraycolsep=0pt
                                      \begin{array}[t]{l}
                                        \Phi(r_2) = c \in \SealableCaps \text{ and } \\
                                        \Phi(r_3) = \seal{\sigma,\_} \text{ and } \\
                                        \vsc = \sealed{c,\sigma}
                                      \end{array}\\
                                      & \\
                                      \failed & \text{otherwise}
                                    \end{cases}
\end{align*}

Unlike CHERI, we want sealed capabilies to be useable in a somewhat normal jump as well as the $\scall{}{}$. We cannot just use $\sjmp{}$ because it only takes one argument and we need to be able to jump to a code/data pair. To this end we introduce $\sxjmp{}{}$:
\begin{align*}
  \ssem{\sxjmp{r_1}{r_2}} = & 
                              \begin{cases}
                                \Phi
                                \arraycolsep=0pt
                                \begin{array}[t]{l}
                                  [\reg.\pcreg \mapsto c_1] \\
                                  [\reg.\rdata \mapsto c_2]
                                \end{array} & 
                                \arraycolsep=0pt
                                \begin{array}[t]{l}
                                  \Phi(r_1) = \sealed{c_1,\sigma} \text{ and } \\
                                  \Phi(r_2) = \sealed{c_2,\sigma} \text{ and } \\
                                  \nonexec{c_2}
                                \end{array}\\
%
                                \\
%
                                \failed & \text{otherwise}
                              \end{cases}
\end{align*}
\lau{This is $\sxjmp{}{}$ for the target language - this one also needs to allow jumps to the return pointer}
Notice, $\rdata$ serves the same function as the IDC register in CHERI, i.e.\ this is where the data capability is put. In the interpretation of $\sxjmp{}{}$, we require $c_2$ to be non-executable because CHERI does it. We are, however, not certain whether this requirement is strictly necessary (maybe it is dangerous to $\sxjmp{r}{r}$ where $r$ points to the same capability? Maybe it can be use to unseal a code capability by first sealing another code capability with the seal, $\sxjmp{}{}$ with the newly sealed and the old sealed capability. Now both capabilities are unsealed). CHERI also checks whether $c_1$ is executable and in range, but we do this on every step of the operational semantics, so we do not repeat that check.

\begin{align*}
  \ssem{\sjmp{r}} = & \dots
\end{align*}

\begin{itemize}
\item $\scall{}{}$ does not do argument spilling anymore. We leave it to the caller to push the arguments for spilling to the stack and update the stack pointer accordingly.
\end{itemize}

@@

\begin{align*}
  \ssem{\sxjmp{r_1}{r_2}} = & 
                              \begin{cases}
                                \Phi
                                \arraycolsep=0pt
                                \begin{array}[t]{l}
                                  [\reg.\pcreg \mapsto c_1] \\
                                  [\reg.\rdata \mapsto c_2]
                                \end{array} & 
                                \arraycolsep=0pt
                                \begin{array}[t]{l}
                                  \Phi(r_1) = \sealed{c_1,\sigma} \text{ and } \\
                                  \Phi(r_2) = \sealed{c_2,\sigma} \text{ and } \\
                                  \nonexec{c_2}
                                \end{array}
\\
\failed & \text{otherwise}
                              \end{cases}
\end{align*}
